---
title: "Using Demor package for basic demographic analysis"
output:
  github_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

The goal of `demor` is to provide you with:\
1. the most basic functions for demographic analysis\
2. some data

## Installation

You can install the development version of `demor` from [GitHub](https://github.com/) with:

``` r

# install.packages("devtools")
devtools::install_github("vadvu/demor")
```

## Get ROSBRIS data

For getting data from [RosBris](http://demogr.nes.ru/index.php/ru/demogr_indicat/data) there is a function `get_rosbris()` that can download data on mortality/fertility by 1/5-year age groups from 1989 to the last available year (in 2023 its 2022).\
Worth noting: downloading and preparing the final file can get some time.\
In the chunk below mortality data for 5-age groups with population in "long" format is loading. For more function description use `?get_rosbris`

```{r}
library(demor)

dbm <- get_rosbris(
  #mortality data
  type = "m",
  #what age group download
  age =  5,
  #to get "long" data
  initial = F,
  #last available year (the name of the downloading file contains years, so for the downloading the last year is required)
  lastyear = 2022
)
```

Lets see the data for Russia in 2010 for males and for total population (both urban and rural)

```{r}
dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]
```

## Mortality

### Life table

Now one can create *life table* based on gotten data for 2010-Russia using `LT()`.\
Note, $a_x$ for age 0 is modeled as in Andreev & Kingkade ([2015](https://doi.org/10.4054/DemRes.2015.33.13)).

```{r}
rus2010 <- dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]

LT(
  age = rus2010$age, 
  sex = "m", 
  #age specific mortality rates
  mx = rus2010$mx)
```

Note, from life table one can compute other *functions* (not just $e_x$ or $l_x$):\
1. $CDR = \frac{1}{e_0}$ or Death rate above some age $x: \frac{1}{e_x}$.\
2. Probability of surviving from age $x$ to age $y$: $p=\frac{l_y}{l_x}$.\
3. Probability that a newborn will die between ages $x$ and $x+n$: $\frac{d_x^n}{l_0}$.\
4. Probability that a newborn will die between ages $x$ and $y$: $\frac{l_x-l_y}{l_0}$.\
5. Life course ratio from age $x$ to $y$ that is the fraction of person-years lived from age $x$ onward: $\frac{T_y}{T_x}$.

### Human Life Indicator (HLI)

A good alternative to the *human development indicator* (HDI) is the *human life indicator* (HLI) proposed by Ghislandi, Sanderson and Scherbov ([2019](https://doi.org/10.1111/padr.12205)). It requires just $m_x$ (and it is based on *life table*). Lets calculate it using our example data:

```{r}
hli(
  age = rus2010$age, 
  sex = "m", 
  mx = rus2010$mx)
```

### Years of Life Lost (YLL)

One of the most popular (and relatively young) measure of *lifespan inequality* is "*years of life lost*" (YLL) proposed by Martinez et al. ([2019](https://doi.org/10.1093/ije/dyy254)). As authors claim, "YLL is a valuable measure for public health surveillance, particularly for quantifying the level and trends of premature mortality, identification of leading causes of premature deaths and monitoring the progress of YLL as a key indicator of population health" (ibid., 1368).   

Authors proposed different metrics of *YLL*:    
1. Absolute number of *YLL*: $$YLL_{x,t,c}=D_{x,t,c}*SLE_x$$ that is calculated for age *x*, time *t* and cause of death *c*.  *YLL* for the whole population is just sum of $YLL_x$. *SLE* is the *standard life expectancy* that is invariant over time, sex and population (it's meaning is straightforward: it is the potential maximum life span of an individual, who is not exposed to avoidable health risks or severe injuries and receives appropriate health services), and $D_x$ is a number of deaths at age *x*. Of course, one can calculate *YLL* not for specific cause *c*, but for overall mortality that is called *all-causes YLL*.    
2. *YLL* as proportion: $$YLL^p_{x,t,c}=YLL_{x,t,c}/YLL_{x,t}$$ that is just *cause specific YLL* divided by *all-causes YLL*.   
3. *YLL* rate: $$YLL^r_{x,t,c}=[YLL_{x,t,c}/P_{x,t}] * 100'000$$ where $P_{x,t}$ is population.    
4. Age-standardized *YLL* rate: $$ASYR_{x,t,c} = \sum_x^\omega{[YLL^r_{x,t,c}*W_x]}$$ where $W_x$ is the standard population weight at age *x*, where $\omega$ is the oldest, closing age (for ex., 85+ or 100+). In other words, it's just direct standardization of $YLL^r_{x,t,c}$.    

Let's calculate all-cause *YLL*, *Yll rate* and *ASYR* using Rosbris data that we have downloaded. 

```{r}
#YLL
yll(rus2010$Dx, type = "yll")

```

```{r}
#YLL rate
yll(rus2010$Dx, type = "yll.r", pop = rus2010$N)

```

For *ASYR* one needs standard population. Let's use 2010 population as standard (note, in this case *ASYR* equals *YLL rate* because we use 2010 mortality).

```{r}
#ASYR
yll(rus2010$Dx, type = "asyr", pop = rus2010$N, w = rus2010$N/sum(rus2010$N))

```

Also one can calculate different *YLL* measures using standards that are provided by `demor` as dataframe. 

``` r
demor::sle_stand
```


### Age decomposition of differences in life expectancies

Also one can do simple *decomposition* between 2 populations. Lets use Russia-2000 as *base population* and Russia-2010 as *compared population*

```{r}
rus2010 <- dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]
rus2000 <- dbm[dbm$year==2000 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]

dec <- decomp(mx1 = rus2000$mx, 
              mx2 = rus2010$mx, 
              sex = "m", 
              age = rus2000$age, 
              method = "andreev")
dec
```

Than let us plot the result of `decomp` using R basic plot.

```{r}
barplot(height=dec$ex12, 
        names=dec$age,
        xlab="Age-groups", 
        ylab="Сontribution to the e0 difference")
```

Or using `ggplot2`

```{r}
library(ggplot2)

ggplot(dec, aes( as.factor(age), ex12))+
  geom_bar(stat = "identity", color = "black", fill = "orange3")+
  theme_minimal()+
  labs(x = "Age-groups", 
       y = "Сontribution to the e0 difference")+
  annotate("text", x = "70", y = 0.5, label = paste0("Total difference in e0: ", sum(dec$ex12)))+
  geom_text(aes(label = ex12), vjust = 1.5, color = "white", size = 3.5)

```


### Age and cause decomposition of differences in life expectancies

Also one can do *decomposition* between 2 populations by *age* and *causes*. Lets use example from Andreev & Shkolnikov [spreadsheet](https://www.demogr.mpg.de/en/publications_databases_6118/publications_1904/mpidr_technical_reports/an_excel_spreadsheet_for_the_decomposition_of_a_difference_between_two_values_of_an_aggregate_demographic_4591) where data for US and England and Wales men mortality by some causes are presented.\
Lets see the data

```{r}
data("mdecompex")
head(mdecompex)
```

For `mdecomp` 2 lists with arrays for 2 population are required.

```{r}
#US men
mx1 <- list(all = mdecompex[mdecompex$cnt=="usa",]$all, 
            neoplasm = mdecompex[mdecompex$cnt=="usa",]$neoplasms, 
            circulatory = mdecompex[mdecompex$cnt=="usa",]$circulatory, 
            respiratory = mdecompex[mdecompex$cnt=="usa",]$respiratory,
            digestive = mdecompex[mdecompex$cnt=="usa",]$digestive,
            accident = mdecompex[mdecompex$cnt=="usa",]$accident,
            other = mdecompex[mdecompex$cnt=="usa",]$other)
#England and Wales men
mx2 <- list(all = mdecompex[mdecompex$cnt=="eng",]$all, 
            neoplasm = mdecompex[mdecompex$cnt=="eng",]$neoplasms, 
            circulatory = mdecompex[mdecompex$cnt=="eng",]$circulatory, 
            respiratory = mdecompex[mdecompex$cnt=="eng",]$respiratory,
            digestive = mdecompex[mdecompex$cnt=="eng",]$digestive,
            accident = mdecompex[mdecompex$cnt=="eng",]$accident,
            other = mdecompex[mdecompex$cnt=="eng",]$other)

decm <- mdecomp(mx1 = mx1, 
              mx2 = mx2, 
              sex = "m", 
              age = unique(mdecompex$age)
              )
head(decm)
```

Than let us plot the result of `mdecomp` using [ggplot2](https://github.com/tidyverse/ggplot2). This requires some data transformations

```{r}
library(ggplot2)

decm_plot <- decm[,c(1,3)]
decm_plot$group = colnames(decm)[3]
colnames(decm_plot)[2]<-"ex12"
for(i in 4:ncol(decm)){
  decm_plot_i <- decm[,c(1,i)]
  decm_plot_i$group = colnames(decm)[i]
  colnames(decm_plot_i)[2]<-"ex12"
  decm_plot <- rbind(decm_plot,decm_plot_i)
  rm(decm_plot_i)
}

for (i in unique(decm_plot$group)){
  decm_plot[decm_plot$group==i,]$group <- paste0(i, " (", round(sum(decm_plot[decm_plot$group==i,]$ex12),2), ")")
}

ggplot(data = decm_plot, aes(x = as.factor(age), y = ex12, fill = group))+
  geom_bar(stat="identity", colour = "black")+
  theme_minimal()+
  scale_fill_brewer(palette="Set1")+
  labs(x = "Age", y = "Contribution to difference in ex", fill = "Cause (contribution):", 
       caption  = paste0("Total difference in ex = ", sum(decm[,2]))
       )
```

### Lee-Carter model

Also in the `demor` there is `leecart()` function that provides users with basic *Lee-Carter model* (proposed by Lee and Carter in 1992 and that now has a lot of extensions) for mortality forecasting:

```{r}
leecart_forecast <- leecart(data = dbm[dbm$code==1100 & dbm$territory=="t" & dbm$sex=="m" & dbm$year %in% 2004:2019,c("year", "age", "mx")], 
                            n = 10, 
                            sex = "m", 
                            concise = T
                            )
head(leecart_forecast)
```

One can plot the results using [ggplot2](https://github.com/tidyverse/ggplot2):

```{r}
ggplot(data = leecart_forecast[leecart_forecast$age=="0",], aes(year, ex))+
  geom_line()+geom_ribbon(aes(ymin = ex_low95, ymax = ex_high95), alpha = 0.3, fill = "red")+
  theme_classic()

ggplot(data = leecart_forecast, aes(as.numeric(age), log10(mx), color = as.factor(year)))+
  geom_line()+geom_ribbon(aes(ymin = log10(mx_low95), ymax = log10(mx_high95)), alpha = 0)+
  theme_classic()
```

### Associated single decrement life table

There is `asdt()` function that calculates *associated single decrement life table* (ASDT) for causes of death (*cause-deleted life table*). In other words, by this function one can answer the question "what will be the life expectancy if there is no mortality from cause i?"\
For example in the `demor` data (as it is easy to guess, taken from Andreev & Shkolnikov [spreadsheet](https://www.demogr.mpg.de/en/publications_databases_6118/publications_1904/mpidr_technical_reports/an_excel_spreadsheet_for_the_decomposition_of_a_difference_between_two_values_of_an_aggregate_demographic_4591)) on mortality of US men in 2002 by some causes is added. Let me show what would be $e_x$ if there is no deaths from neoplasm (i)

```{r}
data("asdtex")

asdt_neoplasm <- asdt(age = asdtex$age, 
                      sex = "m",
                      m_all = asdtex$all, 
                      m_i = asdtex$neoplasms, 
                      full = F, 
                      method = "chiang1968")
asdt_neoplasm[,c("age", "ex", "ex_without_i")]
```

One can plot the results using [ggplot2](https://github.com/tidyverse/ggplot2):

```{r}
library(ggplot2)

ggplot(data = asdt_neoplasm, aes(x = age))+
  geom_line(aes(y = ex, color = "ex"), size = 1)+
  geom_line(aes(y = ex_without_i, color = "ex without neoplasms"), size = 1)+
  theme_minimal()

ggplot(data = asdt_neoplasm, aes(x = age))+
  geom_line(aes(y = lx, color = "ex"), size = 1)+
  geom_line(aes(y = l_not_i, color = "ex without neoplasms"), size = 1)+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()

```

## Fertility

For the analysis of fertility in the `demor` there are only a few (1...) functions, due to the author's preference for mortality analysis...   

### TFR

Lets get basic *fertility data* (asFR or $f_x$) from [RosBris](http://demogr.nes.ru/index.php/ru/demogr_indicat/data) using `get_rosbris()`

```{r}
dbf <- get_rosbris(
  #fertility data
  type = "f",
  #what age group download
  age =  5,
  #to get "long" data
  initial = F,
  #last available year
  lastyear = 2022
)
```

For the example Russia-2010 is gotten

```{r}
rus2010f <- dbf[dbf$year==2010 & dbf$code==1100 & dbf$territory=="t",]
rus2010f
```

Now one can compute *TFR*:

```{r}
tfr(
  #asFR
  rus2010f$fx,
  #age interval
  age.int = 5
    )
```

## Other functions

Also in the `demor` there are some additional functions.\
One of them is `plot_pyr` that plots population pyramid using [ggplot2](https://github.com/tidyverse/ggplot2)\
Lets create population pyramid using midyear population from Rosbris mortality data. We already have data in `dbm`.

```{r}
plot_pyr(
  popm = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="m",]$N,
  popf = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$N, 
  ages = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$age)
```

Also one can redesigned plot using [ggplot2](https://github.com/tidyverse/ggplot2) functions:

```{r}
plot <- 
  plot_pyr(
  popm = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="m",]$N,
  popf = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$N, 
  ages = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$age)

plot + 
  labs(y = "My y", x = "Age")+
  theme_minimal()
```

## References
* Center for Demographic Research, Moscow (Russia). 2023. “Russian Fertility and Mortality Database (Rosbris)” [URL](http://www.demogr.nes.ru/en/demogr_indicat/data)    
* Andreev, E. M., & Kingkade, W. W. (2015). Average age at death in infancy and infant mortality level: Reconsidering the Coale-Demeny formulas at current levels of low mortality. Demographic Research, 33, 363-390.    
* Ghislandi, S., Sanderson, W. C., & Scherbov, S. (2019). A simple measure of human development: The human life indicator. Population and development review, 45(1), 219.   
* Andreev, E. M., & Shkolnikov, V. M. (2012). An Excel spreadsheet for the decomposition of a difference between two values of an aggregate demographic measure by stepwise replacement running from young to old ages. MPIDR Technical Report TR-2012-002. [URL](https://www.demogr.mpg.de/en/publications_databases_6118/publications_1904/mpidr_technical_reports/an_excel_spreadsheet_for_the_decomposition_of_a_difference_between_two_values_of_an_aggregate_demographic_4591)    
* Lee, R. D., & Carter, L. R. (1992). Modeling and forecasting US mortality. Journal of the American statistical association, 87(419), 659-671.    












