---
title: "Get started with `demor`"
output: rmarkdown::html_vignette
toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Get started with `demor`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "#>",
  out.width = "100%"
)
```

# Installation

You can install the development version of `demor` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vadvu/demor")
```

```{r setup, echo=FALSE}
library(demor)
library(ggplot2)
library(dplyr)
```

# Get Rosbris data

For getting data from [RosBris](http://demogr.nes.ru/index.php/ru/demogr_indicat/data) there is a function `get_rosbris()` that can download data on mortality/fertility by 1/5-year age groups from 1989 to the last available year (in 2023 its 2022).\
Worth noting: downloading and preparing the final file can get some time.\
In the chunk below mortality data (`type = "m"`, for fertility see [Fertility section](#get-fertility-data)) for 5-age groups with population in "long" format is loading. For more function description use `?get_rosbris`

```{r}
dbm <- get_rosbris(
  #mortality data
  type = "m",
  #what age group download
  age =  5,
  #to get "long" data
  initial = F,
  #last available year (the name of the downloading file contains years, so for the downloading the last year is required)
  lastyear = 2022
)
```

Lets see the data for Russia in 2010 for males and for total population (both urban and rural)

```{r}
dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]
```

# Mortality

## Life table

Now one can create *life table* based on gotten data for 2010-Russia using `LT()`.\
Note, $a_x$ for age 0 is modeled as in Andreev & Kingkade ([2015](https://doi.org/10.4054/DemRes.2015.33.13)).

```{r}
rus2010 <- dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]

LT(
  age = rus2010$age, 
  sex = "m", 
  #age specific mortality rates
  mx = rus2010$mx)
```

Note, from life table one can compute other *functions* (not just $e_x$ or $l_x$) and interesting things: \    
1. $CDR = \frac{1}{e_0}$ or Death rate above some age $x: 1/e_x$ \
2. Probability of surviving from age $x$ to age $y$: $p=l_y/l_x$ \
3. Probability that a newborn will die between ages $x$ and $x+n$: $d_x^n/l_0$ \
4. Probability that a newborn will die between ages $x$ and $y$: $(l_x-l_y)/l_0$ \
5. Life course ratio from age $x$ to $y$ that is the fraction of person-years lived from age $x$ onward: $T_y/T_x$ \
6. Crude estimate of the number of births needed to “replace” expected deaths: $P/e_0$ where $P$ is total population \    
See also Demography handbook by Preston, Heuveline and Guillot [(2001, p. 38-71)](https://gwern.net/doc/statistics/2001-preston-demography.pdf) 

## Human Life Indicator (HLI)

A good alternative to the *human development indicator* (HDI) is the *human life indicator* (HLI) proposed by Ghislandi, Sanderson and Scherbov ([2019](https://doi.org/10.1111/padr.12205)). It requires just $m_x$ (and it is based on *life table*). It is calculated as geometric mean of lifespans: $$HLI = \prod_x^\omega{(x+a_x)^{d_x}}$$ where *x* is age and *a*, *d* are functions from *life table*.

Lets calculate it using our example data:

```{r}
hli(
  age = rus2010$age, 
  sex = "m", 
  mx = rus2010$mx)
```

## Years of Life Lost (YLL)

One of the most popular (and relatively young) measure of *lifespan inequality* is "*years of life lost*" (YLL) proposed by Martinez et al. ([2019](https://doi.org/10.1093/ije/dyy254)). As authors claim, "YLL is a valuable measure for public health surveillance, particularly for quantifying the level and trends of premature mortality, identification of leading causes of premature deaths and monitoring the progress of YLL as a key indicator of population health" (ibid., 1368).

Authors proposed different metrics of *YLL*:\
1. Absolute number of *YLL*: $$YLL_{x,t,c}=D_{x,t,c}*SLE_x$$ that is calculated for age *x*, time *t* and cause of death *c*. *YLL* for the whole population is just sum of $YLL_x$. *SLE* is the *standard life expectancy* that is invariant over time, sex and population (it's meaning is straightforward: it is the potential maximum life span of an individual, who is not exposed to avoidable health risks or severe injuries and receives appropriate health services), and $D_x$ is a number of deaths at age *x*. Of course, one can calculate *YLL* not for specific cause *c*, but for overall mortality that is called *all-causes YLL*.\
2. *YLL* as proportion: $$YLL^p_{x,t,c}=YLL_{x,t,c}/YLL_{x,t}$$ that is just *cause specific YLL* divided by *all-causes YLL*.\
3. *YLL* rate: $$YLL^r_{x,t,c}=[YLL_{x,t,c}/P_{x,t}] * 100'000$$ where $P_{x,t}$ is population.\
4. Age-standardized *YLL* rate: $$ASYR_{x,t,c} = \sum_x^\omega{[YLL^r_{x,t,c}*W_x]}$$ where $W_x$ is the standard population weight at age *x*, where $\omega$ is the oldest, closing age (for ex., 85+ or 100+). In other words, it's just direct standardization of $YLL^r_{x,t,c}$.

Let's calculate all-cause *YLL*, *Yll rate* and *ASYR* using Rosbris data that we have downloaded.

```{r}
#YLL
yll(rus2010$Dx, type = "yll")

```

```{r}
#YLL rate
yll(rus2010$Dx, type = "yll.r", pop = rus2010$N)

```

For *ASYR* one needs standard population. Let's use 2010 population as standard (note, in this case *ASYR* equals *YLL rate* because we use 2010 mortality).

```{r}
#ASYR
yll(rus2010$Dx, type = "asyr", pop = rus2010$N, w = rus2010$N/sum(rus2010$N))

```

Also one can calculate different *YLL* measures using standards that are provided by `demor` as dataframe.

``` r
demor::sle_stand
```

## Age decomposition of differences in life expectancies

Also one can do simple *decomposition* between 2 populations. Lets use Russia-2000 as *base population* and Russia-2010 as *compared population*

```{r}
rus2010 <- dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]
rus2000 <- dbm[dbm$year==2000 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]

dec <- decomp(mx1 = rus2000$mx, 
              mx2 = rus2010$mx, 
              sex = "m", 
              age = rus2000$age, 
              method = "andreev")
dec
```

Than let us plot the result of `decomp` using R basic plot.

```{r}
barplot(height=dec$ex12, 
        names=dec$age,
        xlab="Age-groups", 
        ylab="Сontribution to the e0 difference")
```

Or using `ggplot2`

```{r}
library(ggplot2)

ggplot(dec, aes( as.factor(age), ex12))+
  geom_bar(stat = "identity", color = "black", fill = "orange3")+
  theme_minimal()+
  labs(x = "Age-groups", 
       y = "Сontribution to the e0 difference")+
  annotate("text", x = "70", y = 0.5, label = paste0("Total difference in e0: ", sum(dec$ex12)))+
  geom_text(aes(label = ex12), vjust = 1.5, color = "white", size = 3.5)

```

## Age and cause decomposition of differences in life expectancies

Also one can do *decomposition* between 2 populations by *age* and *causes*. Lets use example from Andreev & Shkolnikov [spreadsheet](https://www.demogr.mpg.de/en/publications_databases_6118/publications_1904/mpidr_technical_reports/an_excel_spreadsheet_for_the_decomposition_of_a_difference_between_two_values_of_an_aggregate_demographic_4591) where data for US and England and Wales men mortality by some causes are presented.\
Lets see the data

```{r}
data("mdecompex")
head(mdecompex)
```

For `mdecomp` 2 lists with arrays for 2 population are required.

```{r}
#US men
mx1 <- list(all = mdecompex[mdecompex$cnt=="usa",]$all, 
            neoplasm = mdecompex[mdecompex$cnt=="usa",]$neoplasms, 
            circulatory = mdecompex[mdecompex$cnt=="usa",]$circulatory, 
            respiratory = mdecompex[mdecompex$cnt=="usa",]$respiratory,
            digestive = mdecompex[mdecompex$cnt=="usa",]$digestive,
            accident = mdecompex[mdecompex$cnt=="usa",]$accident,
            other = mdecompex[mdecompex$cnt=="usa",]$other)
#England and Wales men
mx2 <- list(all = mdecompex[mdecompex$cnt=="eng",]$all, 
            neoplasm = mdecompex[mdecompex$cnt=="eng",]$neoplasms, 
            circulatory = mdecompex[mdecompex$cnt=="eng",]$circulatory, 
            respiratory = mdecompex[mdecompex$cnt=="eng",]$respiratory,
            digestive = mdecompex[mdecompex$cnt=="eng",]$digestive,
            accident = mdecompex[mdecompex$cnt=="eng",]$accident,
            other = mdecompex[mdecompex$cnt=="eng",]$other)

decm <- mdecomp(mx1 = mx1, 
              mx2 = mx2, 
              sex = "m", 
              age = unique(mdecompex$age)
              )
head(decm)
```

Than let us plot the result of `mdecomp` using [ggplot2](https://github.com/tidyverse/ggplot2). This requires some data transformations

```{r}
library(ggplot2)

decm_plot <- decm[,c(1,3)]
decm_plot$group = colnames(decm)[3]
colnames(decm_plot)[2]<-"ex12"
for(i in 4:ncol(decm)){
  decm_plot_i <- decm[,c(1,i)]
  decm_plot_i$group = colnames(decm)[i]
  colnames(decm_plot_i)[2]<-"ex12"
  decm_plot <- rbind(decm_plot,decm_plot_i)
  rm(decm_plot_i)
}

for (i in unique(decm_plot$group)){
  decm_plot[decm_plot$group==i,]$group <- paste0(i, " (", round(sum(decm_plot[decm_plot$group==i,]$ex12),2), ")")
}

ggplot(data = decm_plot, aes(x = as.factor(age), y = ex12, fill = group))+
  geom_bar(stat="identity", colour = "black")+
  theme_minimal()+
  scale_fill_brewer(palette="Set1")+
  labs(x = "Age", y = "Contribution to difference in ex", fill = "Cause (contribution):", 
       caption  = paste0("Total difference in ex = ", sum(decm[,2]))
       )
```

## Associated single decrement life table

There is `asdt()` function that calculates *associated single decrement life table* (ASDT) for causes of death (*cause-deleted life table*). In other words, by this function one can answer the question "what will be the life expectancy if there is no mortality from cause i?"\
For example in the `demor` data (as it is easy to guess, taken from Andreev & Shkolnikov [spreadsheet](https://www.demogr.mpg.de/en/publications_databases_6118/publications_1904/mpidr_technical_reports/an_excel_spreadsheet_for_the_decomposition_of_a_difference_between_two_values_of_an_aggregate_demographic_4591)) on mortality of US men in 2002 by some causes is added. Let me show what would be $e_x$ if there is no deaths from neoplasm (i)

```{r}
data("asdtex")

asdt_neoplasm <- asdt(age = asdtex$age, 
                      sex = "m",
                      m_all = asdtex$all, 
                      m_i = asdtex$neoplasms, 
                      full = F, 
                      method = "chiang1968")
asdt_neoplasm[,c("age", "ex", "ex_without_i")]

```

One can plot the results using [ggplot2](https://github.com/tidyverse/ggplot2):

```{r}
library(ggplot2)

ggplot(data = asdt_neoplasm, aes(x = age))+
  geom_line(aes(y = ex, color = "ex"), size = 1)+
  geom_line(aes(y = ex_without_i, color = "ex without neoplasms"), size = 1)+
  theme_minimal()

ggplot(data = asdt_neoplasm, aes(x = age))+
  geom_line(aes(y = lx, color = "ex"), size = 1)+
  geom_line(aes(y = l_not_i, color = "ex without neoplasms"), size = 1)+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  theme_classic()

```

## Mortality models for mx approximation

In `demor` there is a function `mort.approx` for modeling mx. Now "Gompertz" and "Brass" are supported (see Preston, Heuveline & Guillot, 2001 for more details on the functions). 

Function returns list with estimated model (parameters, R-squred, variance-covariance matrix of parameters) and dataframe with predicted mx.

The example below shows the model that try to approximate mx of russian men in 2010 using Brass function and rissian mortality of 2000 as standard mortality. 

```{r}
rus2010 <- dbm[dbm$year==2010 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]
rus2000 <- dbm[dbm$year==2000 & dbm$code==1100 & dbm$sex=="m" & dbm$territory=="t",]


brass_2010 <- mort.approx(mx = rus2010$mx,
                          age = rus2010$age,
                          model = "Brass",
                          standard.mx = rus2000$mx,
                          boot = T)
brass_2010[[1]]
```

Lets plot the modeled and observed mx with bootstrapped 95% CI.

```{r}
brass_2010[[2]] %>% 
  mutate(mx = rus2010$mx) %>% 
  ggplot(aes(x = age))+
  geom_line(aes(y = log(mx), color = "Observed"), linewidth = 1)+
  geom_line(aes(y = log(mx.pred), color = "Predicted"), linewidth = 1)+
  geom_ribbon(aes(ymin = log(low_2.5prc), ymax = log(high_97.5prc)), 
              alpha = 0, color = "red", linetype = "dashed")+
  theme_classic()+
  scale_color_manual(values = c("blue", "red"))+
  labs(x = "Age", y = "ln mx")
```



# Fertility

For the analysis of fertility in the `demor` there are only a few functions, due to the author's preference for mortality analysis...

## Get fertility data

Lets get basic *fertility data* (asFR or $f_x$) from [RosBris](http://demogr.nes.ru/index.php/ru/demogr_indicat/data) using `get_rosbris()`

```{r}
dbf <- get_rosbris(
  #fertility data
  type = "f",
  #what age group download
  age =  1,
  #to get "long" data
  initial = F,
  #last available year
  lastyear = 2022
)
```

For the example Russia-2010 is gotten

```{r}
rus2010f <- dbf[dbf$year==2010 & dbf$code==1100 & dbf$territory=="t",]
rus2010f
```
## TFR

Now one can compute total fertility age (*TFR*) - the most popular measure of fertility - that is standardized measure of "the average number of children a woman would bear if she survived through the end of the reproductive age span and experienced at each age a particular set of age-specific fertility ages" [Preston et al, 2001, p. 95](https://gwern.net/doc/statistics/2001-preston-demography.pdf).

```{r}
tfr(
  #asFR
  rus2010f$fx,
  #age interval
  age.int = 1
    )
```

## MAC

MAC is a mean age at childbearing. One can compute it using $f_x$

```{r}
mac(
  #asFR
  rus2010f$fx,
  #array with ages
  ages = 15:55
    )
```

## Fertility models for ASFR approximation

In `demor` there is a function `fert.approx` for modeling ASFR. Now "Hadwiger" and "Gamma" are supported (see Peristera & Kostaki, 2007 for more details on the functions). 

Hadwiger model is as follows: $$f(age) = \frac{ab}{c} \frac{c}{age}^{3/2} exp[-b^2(\frac{c}{age}+\frac{age}{c}-2)]$$

Gamma model is as follows: $$f(age) = \frac{R}{\Gamma(b)c^b}(age-d)^{b-1} exp[-(\frac{age-d}{c})]$$

Function returns list with estimated model (parameters, R-squred, variance-covariance matrix of parameters) and dataframe with predicted ASFR.

The example below shows the model that try to approximate ASFR of russian women in 2010 using Gamma function. 

```{r}
approximation_2010 = fert.approx(fx = rus2010f$fx, age = 15:55, model = "Gamma", boot = T)
approximation_2010[[1]]
```

Lets plot the modeled and observed ASFR with bootstrapped 95% CI. One can see that model perfectly approximates real ASFR from 15 to 40 ages, while after the fit is not really good. Meanwhile, the model explains 99.7% of variance in observed ASFR. 

```{r}
data = approximation_2010[[2]]
ggplot()+
  theme_bw()+
  geom_line(data = data, aes(x = age, y = predicted, color = "fx from Gamma model"), linewidth = 1)+
  geom_ribbon(data = data,
              aes(ymin = low_2.5prc, ymax = high_97.5prc, x = age, color = "95% CI"), alpha = 0, linetype = "dashed")+
  geom_line(data = data.frame(real = rus2010f$fx, age = 15:55), aes(x = age, y = real, color = "Observed fx"), linewidth = 1)+
  scale_color_manual(values = c("red", "red", "blue"))+
  labs(y = "fx", x = "Age")
```



# Projections

## Lee-Carter model

In the `demor` there is `leecart()` function that provides users with basic *Lee-Carter model* (proposed by Lee and Carter in 1992 and that now has a lot of extensions) for mortality forecasting:

```{r}
leecart_forecast <- leecart(data = dbm[dbm$code==1100 & dbm$territory=="t" & dbm$sex=="m" & dbm$year %in% 2004:2019,c("year", "age", "mx")], 
                            n = 5, 
                            sex = "m", 
                            concise = T
                            )
head(leecart_forecast)
```

One can plot the results using [ggplot2](https://github.com/tidyverse/ggplot2) to comapare predicted data with actual that, however, requires some data handling:

```{r}
#LE data calculation
ledata <- data.frame(year = 2010:2022, lem = NA, lem.predicted = NA, ci.low = NA, ci.high = NA)
for(i in 2010:2022){
  ledata[ledata$year==i,2] <- LT(age = unique(dbm$age), 
                                 sex = "m", 
                                 mx = dbm[dbm$year==i & 
                                            dbm$territory == "t" & 
                                            dbm$code == 1100 & 
                                            dbm$sex == "m",]$mx)[1,"ex"]
  if(i>=2019){
    if(i==2019){
      ledata[ledata$year==i,3:5] <- rep(ledata[ledata$year==i,2],3)
    }else{
      ledata[ledata$year==i,3:5] <- leecart_forecast[leecart_forecast$age=="0" & 
                                                     leecart_forecast$year == i,6:8] %>% as.numeric()
    }
  }
}

ggplot(data = ledata, aes(x = year))+
  geom_point(aes(y = lem, color = "Observed data"))+
  geom_line(aes(y = lem, color = "Observed data"))+
  geom_point(aes(y = lem.predicted, color = "Predicted data"))+
  geom_line(aes(y = lem.predicted, color = "Predicted data"))+
  geom_ribbon(aes(ymin = ci.low, ymax = ci.high), fill = "pink", alpha = 0.5)+
  geom_vline(xintercept = 2019, linetype = "dashed", color = "darkgrey", size = 1)+
  scale_x_continuous(breaks = 2010:2022)+
  scale_y_continuous(breaks = 63:70)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  scale_color_manual(values = c("black", "darkred"))+
  labs(x="Year", y = "Male's LE at birth", colour = "")
```

## Cohort-component model

Soon


# Other functions

## Population pyramid

`plot_pyr` plots population pyramid using [ggplot2](https://github.com/tidyverse/ggplot2)\
Lets create population pyramid using midyear population from Rosbris mortality data. We already have data in `dbm`.

```{r}
plot_pyr(
  popm = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="m",]$N,
  popf = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$N, 
  ages = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$age)
```

Also one can redesigned plot using [ggplot2](https://github.com/tidyverse/ggplot2) functions:

```{r}
plot <- 
  plot_pyr(
  popm = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="m",]$N,
  popf = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$N, 
  ages = dbm[dbm$year==2010 & dbm$code==1100 & dbm$territory=="t" & dbm$sex=="f",]$age)

plot + 
  labs(y = "My y", x = "Age")+
  theme_minimal()
```

## Recoding age

`ages` converts continuous ages to age groups. For example, array with $\{ 0,1,2,3...100 \}$ it can convert to $\{<5,6-9,10-14 ... 85+ \}$ using user-specific lower boundaries.

```{r}
groups <- c(0, 1, 5, 10, 20, 45, 85) #lower boundaries of age groups
age <- sample(0:100, 101) #simulated array with ages

new.age <- ages(x = age, groups = groups, char = TRUE)

table(new.age) #see new groups with the number of values in the group 
```


# References

-   Center for Demographic Research (2023). *Russian Fertility and Mortality Database (Rosbris)*. [URL](http://www.demogr.nes.ru/en/demogr_indicat/data)\
-   Preston, S. H., Heuveline, P., & Guillot, M. (2001). Demography: Measuring and modeling population processes. Blackwell Publishers. ([pdf](https://gwern.net/doc/statistics/2001-preston-demography.pdf))\
-   Andreev, E. M., & Kingkade, W. W. (2015). Average age at death in infancy and infant mortality level: Reconsidering the Coale-Demeny formulas at current levels of low mortality. *Demographic Research*, *33*, 363-390.\
-   Ghislandi, S., Sanderson, W. C., & Scherbov, S. (2019). A simple measure of human development: The human life indicator. *Population and development review*, *45*(*1*), 219-233.\
-   Andreev, E. M., & Shkolnikov, V. M. (2012). *An Excel spreadsheet for the decomposition of a difference between two values of an aggregate demographic measure by stepwise replacement running from young to old ages*. MPIDR Technical Report TR-2012-002. [URL](https://www.demogr.mpg.de/en/publications_databases_6118/publications_1904/mpidr_technical_reports/an_excel_spreadsheet_for_the_decomposition_of_a_difference_between_two_values_of_an_aggregate_demographic_4591)\
-   Lee, R. D., & Carter, L. R. (1992). Modeling and forecasting US mortality. *Journal of the American statistical association*, *87*(*419*), 659-671.\
-   Ghislandi, S., Sanderson, W.C., & Scherbov, S. (2019). A Simple Measure of Human Development: The Human Life Indicator. *Population and Development Review*, *45*, 219--233.\
-   Peristera, P., & Kostaki, A. (2007). Modeling fertility in modern populations. *Demographic Research*, *16*, 141-194.\
